A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

### 2020-05-11 Manuscript/Manuscript-7.tex
vsn 7

Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. In this study, we evaluated the
efficacy of one commonly used parameter (Pagel's \(\lambda\)) to
estimate the strength of phylogenetic signal in phenotypic traits, and
evaluate the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal. We find that the precision of \(\lambda\)
in estimating actual levels of phylogenetic signal is often inaccurate,
and that biological interpretations of the strength of phylogenetic
signal based on \(\lambda\) are therefore compromised. We then propose a
standardized effect size based on \(\kappa\) (\(Z_\kappa\)), which
measures the strength of phylogenetic signal, and places it on a common
scale for statistical comparison. Tests based on \(Z_\kappa\) provide a
mechanism for formally comparing the strength of phylogenetic signal
across datasets, in much the same manner as effect sizes may be used to
summarize patterns in quantitative meta-analysis. Our approach extends
the phylogenetic comparative toolkit to address hypotheses that compare
the strength of phylogenetic signal between various phenotypic traits,
even when those traits are found in different evolutionary lineages or
have different units or scale.

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(\kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (e.g., Münkemüller et al. 2012;
Pavoine and Ricotta 2012; Diniz-Filho et al. 2012; Adams 2014a;
Molina-Venegas and Rodríguez 2017; see also Revell et al. 2008; Revell
2010). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). The (\(\lambda\)) parameter transforms the lengths of the
internal branches of the phylogeny to improve a fit of data to the
phylogeny via maximum likelihood (Pagel 1999; Freckleton et al. 2002).
Pagel's \(\lambda\) ranges from \(0\to1\), with larger values signifying
a greater dependence of observed trait variation on the phylogeny.
Pagel's \(\lambda\) also has the appeal that it may be included in
phylogenetic generalized least-squares regression (PGLS) to account for
the degree of phylogenetic signal in comparative analyses (see
Freckleton et al. 2002). \hfill\break

In addition to functioning as a parameter that is tuned for appropriate
analysis, \(\lambda\) can function as a descriptive statistic to
describe the relative strength of phylogenetic signal in phenotypic
traits, to determine the extent to which shared evolutionary history has
influenced trait covariation among taxa. The appeal of \(\lambda\) as a
descriptive statistic for evolutionary biologists is a basis for
interpreting ``weak'' versus ``strong'' phylogenetic signal; i.e., small
versus large values of \(\lambda\), respectively, in a comparative sense
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Indeed, statements regarding the strength of phylogenetic signal based
on \(\lambda\) are rather common in the evolutionary literature. For
instance, of the 204 papers published in 2019 that estimated and
reported Pagel's \(\lambda\) (found from a literature survey we
conducted in Google.scholar), 40\% interpreted the strength of
phylogenetic signal for at least one phenotypic trait. Further, because
nearly half of the 1,572 \(\lambda\) values reported were near 0 or 1
(Figure 1) where the biological interpretation of \(\lambda\) is known,
this percentage is even higher. \hfill\break

@@

Various other approaches use \(\lambda\) as a parameter that can be
varied for inferences akin to sensitivity analysis. For instance, some
have performed likelihood ratio tests that compare observed model fits
to those obtained when \(\lambda=0\) or \(\lambda=1\) (Freckleton et al.
2002; Cooper et al. 2010; Bose et al. 2019) or evaluated whether
observed \(\lambda\) differs from an expected \(\lambda\), based on
confidence intervals generated for the expected value (Vandelook et al.
2019). Qualitative comparisons of \(\lambda\) estimates have also been
performed for multiple traits on the same phylogenetic tree to infer
whether the strength of phylogenetic signal is greater in one trait as
compared to another (e.g., Liu et al. 2019; Bai et al. 2019).
\hfill\break

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also implies that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we lack a
general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure of the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and descriptive statistics are
effect sizes, and thus many summary measures must first be converted to
statistics with standardized units (i.e., conversion to an effect size)
for meaningful comparison (see Rosenthal 1994). As a consequence, it
follows that only if \(\lambda\) is a statistical effect size can
comparisons of estimates across datasets be interpretable. For the case
of \(\lambda\), this has not yet been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable indicator of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets, and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(\kappa\). Through simulations we
find that the precision of effect sizes based on \(\lambda\)
(\(Z_{\lambda}\)) are less reliable than that those based on \(\kappa\)
(\(Z_\kappa\)), implying that \(Z_\kappa\) is a more robust effect size
measure. We also propose a two-sample test statistic that may be used to
compare the strength of phylogenetic signal among datasets, and provide
an empirical example to demonstrate its use. We conclude that estimates
of phylogenetic signal using Pagel's \(\lambda\) are often inaccurate,
and thus interpreting strength of phylogenetic signal in phenotypic
datasets based on this measure is compromised. By contrast, effect sizes
obtained from \(\kappa\) hold promise for characterizing phylogenetic
signal, and for comparing the strength of phylogenetic signal across
datasets.

@@

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression and ANOVA, the results of our simulations were quite
similar. Regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3; see also Supporting Information), and the
spread of \(\lambda_{est}\) was similar to that observed when
\(\lambda\) was estimated for only the dependent variable, as in Figure
2. Taken together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

To evaluate the utility of \(Z_\kappa\) and \(Z_\lambda\) we calculated
both effect sizes for the simulated datasets generated above, and
summarized the precision of each using its variance
(\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\), Figure 4:
additional results in the Supporting Information). Here two things are
evident. First, estimates of \(Z_\kappa\) linearly track the input
phylogenetic signal whereas estimates of \(Z_\lambda\) do not (Figure
4A, B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_\kappa\). Second, the precision of \(Z_\kappa\) is considerably more
stable as compared with \(Z_\lambda\). This may be seen by calculating
the coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\)) across input
levels of phylogenetic signal. Coefficients of variation in the
precision of \(Z_\kappa\) were up to an order of magnitude smaller for
than for \(Z_\lambda\) (Figure 4C), implying that estimates of the
strength of phylogenetic signal were more reliable and robust when using
\(Z_\kappa\). \hfill\break

@@

To demonstrate the utility of \(\hat{Z}_{12}\) we quantified and
compared the strength of phylogenetic signal of two phenotypic traits
across species of plethodontid salamander. The data were part of a
series of studies examining macroevolutionary trends in phenotypic
diversification in this group (Baken and Adams 2019; Baken et al. 2020).
Our dataset contained surface area to volume ratios (SA:V) and relative
body width (\(\frac{BW}{SVL}\)) for 305 species (Figure 5A). For SA:V,
11 linear body measurements from 2,781 individuals were taken, from
which estimates of the surface area and volume of the head, body, and
tail were calculated and subsequently combined to arrive at the SA:V for
each indiviual (for mathematical details see Baken et al. 2020). Species
means were then obtained. Likewise, body size (SVL) and body width (BW)
measurements were taken from 3,371 individuals, and species means of
relative body width (\(\frac{BW}{SVL}\)) were calculated (data from
Baken and Adams 2019). A time-dated molecular phylogeny for the group
(Bonett and Blair 2017) was then pruned to match the species in the
dataset, resulting in a phylogeny and corresponding phenotypic dataset
containing 305 species. The phylogenetic signal in each trait was then
characterized using \(\kappa\), which was converted to its effect size
(\(Z_\kappa\)) using \texttt{geomorph} 3.2.1 (Adams and Otárola-Castillo
2013; Adams et al. 2020). Finally, the strength of phylogenetic signal
was compared across traits using \(\hat{Z}_{12}\) as described above
(\textbf{to be incorporated in \texttt{geomorph} upon manuscript
acceptance}). \hfill\break

@@

As an alternative, we described a standardized effect size (\(Z\)) for
assessing the strength of phylogenetic signal. \(Z\) expresses the
magnitude of phylogenetic signal as a standard normal deviate, which is
easily interpretable as the strength of phylogenetic signal relative to
the mean. We applied this concept to both \(\lambda\) and \(\kappa\),
and found that \(Z_\kappa\) was a better estimate of the strength of
phylogenetic signal in phenotypic data. First, \(Z_\kappa\) was more
precise than \(Z_{\lambda}\), and precision was more consistent across
the range of input levels of phylogenetic signal. Additionally, values
of \(Z_\kappa\) more accurately tracked known levels of phylogenetic
signal, with changes in the actual strength of phylogenetic signal
reflected in a more linear fashion by concomitant changes in the values
of \(Z_\kappa\). Thus, \(Z_\kappa\) holds promise as a measure of the
relative strength of phylogenetic signal that reflects the magnitude of
this effect in phenotypic data. We therefore recommend that future
studies interested in the strength of phylogenetic signal incorporate
\(Z_\kappa\) as a statistical measure of this effect. \hfill\break

Based on the effect size \(Z_\kappa\), we then proposed a two-sample
test, which provides means of determining whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared to
another, via a hypothesis test. Prior studies have summarized patterns
of variation in phylogenetic signal across datasets using summary test
values, such as \(\kappa\) (e.g., Blomberg et al. 2003). However,
\(\kappa\) does not scale linearly with input levels of phylogenetic
signal, and its variance increases (i.e., precision decreases) with
increasing strength of phylogenetic signal (Münkemüller et al. 2012;
Diniz-Filho et al. 2012: see also Supporting Information). Thus,
\(\kappa\) should not be considered an effect size that measures the
strength of phylogenetic signal on a common scale. By contrast,
standardizing \(\kappa\) (\(Z_\kappa\), via equation 2) alleviates these
concerns, and facilitates formal statistical comparisons of the strength
of signal across datasets. Thus when viewed from this perspective, the
approach developed here aligns well with other statistical approaches
such as meta-analysis (sensu Hedges and Olkin 1985; Glass 1976; Arnqvist
and Wooster 1995), where summary statistics across datasets are
converted to standardized effect sizes for subsequent ``higher order''
statistical summaries or comparisons. As such our approach enables
evolutionary biologists to quantitatively examine the relative strength
of phylogenetic signal across a wide range of phenotypic traits, and
thus opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.
\hfill\break

One important advantage of the approach advocated here is that the
resulting effect sizes (\(Z_\kappa\)) are dimensionless, as the units of
measurement cancel out during the calculation of \(Z\) (Sokal and Rohlf
2012). Thus, \(Z_\kappa\) represents the strength of phylogenetic signal
on a common and comparable scale -- measured in standard deviation --
regardless of the initial units and original scale of the phenotypic
variables under investigation. This means that the strength of
phylogenetic signal may be compared across datsets for continuous
phenotypic traits measured in different units and scale, because those
units have been standardized through their conversion to \(Z_\kappa\).
For example, our approach could be utilized to determine whether the
strength of phylogenetic signal (say, in response to ecological
differentiation) is stronger in morphological traits (linear traits:
\(mm\)), physiological traits (metabolic rate: \(\frac{O^2}{min}\)), or
behavioral traits (aggression: \(\frac{\#{displays}}{second}\)). In
fact, our empirical example provided such a comparison, as SA:V is
represented in \(mm^{-1}\) while relative body size is a unitless ratio
(\(\frac{BW}{SVL}\)). Additionally, our method is capable of comparing
the strength of phylogenetic signal in traits of different
dimensionality, as estimates of phylogenetic signal using \(\kappa\)
have been generalized for multivariate data (\(K_{mult}\): see Adams
2014a). Furthermore, tests based on \(\hat{Z}_{12}\) may be utilized for
comparing the strength of phylogenetic signal among datasets containing
a different number of species, and even for phenotypes obtained from
species in different lineages, because their phylogenetic
non-independence and observed variation are taken into account in the
generation of the empirical sampling distribution via permutation.
\hfill\break

>>>>

Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. In this study, we evaluated the
efficacy of one commonly used parameter (Pagel's \(\lambda\)) to
estimate the strength of phylogenetic signal in phenotypic traits, and
evaluate the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal. We find that the precision of \(\lambda\)
in estimating actual levels of phylogenetic signal is often inaccurate,
and that biological interpretations of the strength of phylogenetic
signal based on \(\lambda\) are therefore compromised. We then propose a
standardized effect size based on \(\kappa\) (\(Z_\kappa\)), which
measures the strength of phylogenetic signal, and places it on a common
scale for statistical comparison. Tests based on \(Z_\kappa\) provide a
mechanism for formally comparing the strength of phylogenetic signal
across datasets, in much the same manner as effect sizes may be used to
summarize patterns in quantitative meta-analysis. Our approach extends
the phylogenetic comparative toolkit to address hypotheses that compare
the strength of phylogenetic signal between various phenotypic traits,
even when those traits are found in different evolutionary lineages or
have different units or scales.

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(\kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (e.g., Münkemüller et al. 2012;
Pavoine and Ricotta 2012; Diniz-Filho et al. 2012; Adams 2014a;
Molina-Venegas and Rodríguez 2017; see also Revell et al. 2008; Revell
2010). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). The parameter (\(\lambda\)) transforms the lengths of the
internal branches of the phylogeny to improve the fit of data to the
phylogeny via maximum likelihood (Pagel 1999; Freckleton et al. 2002).
Pagel's \(\lambda\) ranges from \(0\to1\), with larger values signifying
a greater dependence of observed trait variation on the phylogeny.
Pagel's \(\lambda\) also has the appeal that it may be included in
phylogenetic generalized least-squares regression (PGLS) to account for
the degree of phylogenetic signal in comparative analyses (see
Freckleton et al. 2002). \hfill\break

In addition to functioning as a parameter that is tuned for appropriate
analysis, \(\lambda\) can function as a descriptive statistic
characterizing the relative strength of phylogenetic signal in
phenotypic traits to determine the extent to which shared evolutionary
history has influenced trait covariation among taxa. The appeal of
\(\lambda\) as a descriptive statistic is that it serves as the basis
for interpreting ``weak'' versus ``strong'' phylogenetic signal; i.e.,
small versus large values of \(\lambda\), respectively, in a comparative
sense (e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al.
2019). Indeed, statements regarding the strength of phylogenetic signal
based on \(\lambda\) are rather common in the evolutionary literature.
For instance, of the 204 papers published in 2019 that estimated and
reported Pagel's \(\lambda\) (found from a literature survey we
conducted in Google.scholar), 40\% explicitly interpreted the strength
of phylogenetic signal for at least one phenotypic trait. Further,
because nearly half of the 1,572 \(\lambda\) values reported were near 0
or 1 (Figure 1) where the biological interpretation of \(\lambda\) is
known (\(\lambda=0\) represents no phylogenetic signal, while
\(\lambda=1\) is phylogenetic signal as expected under Brownian motion),
this percentage is even higher. \hfill\break

@@

Various other approaches use \(\lambda\) as a parameter that can be
adjusted for inferences that are akin to sensitivity analysis. For
instance, some have performed likelihood ratio tests that compare
observed model fits to those obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019) or evaluated whether observed \(\lambda\) differs from an expected
\(\lambda\), based on confidence intervals generated for the expected
value (Vandelook et al. 2019). Qualitative comparisons of \(\lambda\)
estimates have also been performed for multiple traits on the same
phylogenetic tree to infer whether the strength of phylogenetic signal
is greater in one trait as compared to another (e.g., Liu et al. 2019;
Bai et al. 2019). \hfill\break

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. However, equating values of \(\lambda\) directly
to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., low number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also implies that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we lack a
general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure of the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunately, not all model parameters and descriptive statistics are
effect sizes, and thus many summary measures must first be converted to
statistics with standardized units (i.e., conversion to an effect size)
for meaningful comparison (see Rosenthal 1994). As a consequence, it
follows that only if \(\lambda\) is a statistical effect size can
comparisons of estimates across datasets be interpretable. However, the
calculation and statisticl behavior of \(\lambda\) as an effect size has
not yet been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable indicator of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(\kappa\). Through simulations we
find that the precision of effect sizes based on \(\lambda\)
(\(Z_{\lambda}\)) are less reliable than that those based on \(\kappa\)
(\(Z_\kappa\)), implying that \(Z_\kappa\) is a more robust effect size
measure. We also propose a two-sample test statistic that may be used to
compare the strength of phylogenetic signal among datasets, and provide
an empirical example to demonstrate its use. We conclude that estimates
of phylogenetic signal using Pagel's \(\lambda\) are often inaccurate,
and thus interpreting strength of phylogenetic signal in phenotypic
datasets based on this measure is compromised. By contrast, effect sizes
obtained from \(\kappa\) hold promise for characterizing phylogenetic
signal, and for comparing the strength of phylogenetic signal across
datasets.

@@

When phylogenetic signal was investigated on balanced or pectinate
trees, patterns in the precision of \(\lambda\) were largely the same,
with decreased precision at intermediate levels of phylogenetic signal
(Supporting Information). Likewise, when \(\lambda\) was co-estimated
with regression parameters in PGLS regression and ANOVA, the results of
our simulations were quite similar. Regression parameters (\(\beta\))
were accurately estimated, confirming earlier findings of Revell (2010)
(see Supporting Information). However, estimates of phylogenetic signal
(\(\lambda\)) were less precise (Figure 3; see also Supporting
Information), and the spread of \(\lambda_{est}\) was similar to that
observed when \(\lambda\) was estimated for only the dependent variable,
as in Figure 2. Taken together, these findings reveal that
\(\lambda_{est}\) does not precisely characterize known levels of
phylogenetic signal in phenotypic datasets, and that biological
interpretations of the strength of phylogenetic signal based on
\(\lambda\) may be highly inaccurate. \hfill\break

@@

To evaluate the utility of \(Z_\kappa\) and \(Z_\lambda\) we calculated
both effect sizes for the simulated datasets generated above, and
summarized the precision of each using its variance
(\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\), Figure 4:
additional results in the Supporting Information). Here two things are
evident. First, estimates of \(Z_\kappa\) linearly track the input
phylogenetic signal whereas estimates of \(Z_\lambda\) do not (Figure
4A, B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_\kappa\). Second, the precision of \(Z_\kappa\) is considerably more
stable as compared with \(Z_\lambda\). This may be seen by calculating
the coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\)) across input
levels of phylogenetic signal. Coefficients of variation in the
precision of \(Z_\kappa\) were up to an order of magnitude smaller than
for \(Z_\lambda\) (Figure 4C), implying that estimates of the strength
of phylogenetic signal were more reliable and robust when using
\(Z_\kappa\). \hfill\break

@@

To demonstrate the utility of \(\hat{Z}_{12}\) we quantified and
compared the strength of phylogenetic signal of two phenotypic traits
across species of plethodontid salamander. The data were part of a
series of studies examining macroevolutionary trends in phenotypic
diversification in this group (Baken and Adams 2019; Baken et al. 2020).
Our dataset contained surface area to volume ratios (SA:V) and relative
body width (\(\frac{BW}{SVL}\)) for 305 species (Figure 5A). For SA:V,
11 linear body measurements from 2,781 individuals were taken, from
which estimates of the surface area and volume of the head, body, and
tail were calculated and subsequently combined to arrive at the SA:V for
each indiviual (for mathematical details see Baken et al. 2020). Species
means were then obtained. Likewise, body size (SVL) and body width (BW)
measurements were taken from 3,371 individuals, and species means of
relative body width (\(\frac{BW}{SVL}\)) were calculated (data from
Baken and Adams 2019). A time-dated molecular phylogeny for the group
(Bonett and Blair 2017) was then pruned to match the species in the
phenotypic dataset. The phylogenetic signal in each trait was then
characterized using \(\kappa\), which was converted to its effect size
(\(Z_\kappa\)) using \texttt{geomorph} 3.2.1 (Adams and Otárola-Castillo
2013; Adams et al. 2020). Finally, the strength of phylogenetic signal
was compared across traits using \(\hat{Z}_{12}\) as described above
(\textbf{to be incorporated in \texttt{geomorph} upon manuscript
acceptance}). \hfill\break

@@

As an alternative, we described a standardized effect size (\(Z\)) for
assessing the strength of phylogenetic signal. \(Z\) expresses the
magnitude of phylogenetic signal as a standard normal deviate, which is
easily interpretable as the strength of phylogenetic signal relative to
the mean. We applied this concept to both \(\lambda\) and \(\kappa\),
and found that \(Z_\kappa\) was a better estimate of the strength of
phylogenetic signal in phenotypic data. First, \(Z_\kappa\) was more
precise than \(Z_\lambda\), and precision was more consistent across the
range of input levels of phylogenetic signal. Additionally, values of
\(Z_\kappa\) more accurately tracked known changes in the magnitude of
phylogenetic signal, as demonstrated by the linear relationship between
\(Z_\kappa\) and \(\lambda_{in}\). Thus, \(Z_\kappa\) holds promise as a
measure of the relative strength of phylogenetic signal that reflects
the magnitude of this effect in phenotypic data. We therefore recommend
that future studies interested in the strength of phylogenetic signal
incorporate \(Z_\kappa\) as a statistical measure of this effect.
\hfill\break

Based on the effect size \(Z_\kappa\), we then proposed a two-sample
test, which provides means of determining whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared to
another, via a hypothesis test. Prior studies have summarized patterns
of variation in phylogenetic signal across datasets using summary test
values, such as \(\kappa\) (e.g., Blomberg et al. 2003). However,
\(\kappa\) does not scale linearly with input levels of phylogenetic
signal, and its variance increases (i.e., precision decreases) with
increasing strength of phylogenetic signal (Münkemüller et al. 2012;
Diniz-Filho et al. 2012: see also Supporting Information). Thus,
\(\kappa\) should not be considered an effect size that measures the
strength of phylogenetic signal on a common scale. By contrast,
standardizing \(\kappa\) (\(Z_\kappa\), via equation 2) alleviates these
concerns, and facilitates formal statistical comparisons of the strength
of signal across datasets. Thus when viewed from this perspective, the
approach developed here aligns well with other statistical approaches
such as meta-analysis (sensu Hedges and Olkin 1985; Glass 1976; Arnqvist
and Wooster 1995), where summary statistics across datasets are
converted to standardized effect sizes for subsequent ``higher order''
statistical summaries or comparisons. As such, our approach enables
evolutionary biologists to quantitatively examine the relative strength
of phylogenetic signal across a wide range of phenotypic traits, and
thus opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.
\hfill\break

One important advantage of the approach advocated here is that the
resulting effect sizes (\(Z_\kappa\)) are dimensionless, as the units of
measurement cancel out during the calculation of \(Z\) (Sokal and Rohlf
2012). Thus, \(Z_\kappa\) represents the strength of phylogenetic signal
on a common and comparable scale -- measured in standard deviations --
regardless of the initial units and original scale of the phenotypic
variables under investigation. This means that the strength of
phylogenetic signal may be compared across datsets for continuous
phenotypic traits measured in different units and scale, because those
units have been standardized through their conversion to \(Z_\kappa\).
For example, our approach could be utilized to determine whether the
strength of phylogenetic signal (say, in response to ecological
differentiation) is stronger in morphological traits (linear traits:
\(mm\)), physiological traits (metabolic rate: \(\frac{O^2}{min}\)), or
behavioral traits (aggression: \(\frac{\#{displays}}{second}\)). In
fact, our empirical example provided such a comparison, as SA:V is
represented in \(mm^{-1}\) while relative body size is a unitless ratio
(\(\frac{BW}{SVL}\)). Additionally, our method is capable of comparing
the strength of phylogenetic signal in traits of different
dimensionality, as estimates of phylogenetic signal using \(\kappa\)
have been generalized for multivariate data (\(\kappa_{mult}\): see
Adams 2014a). Furthermore, tests based on \(\hat{Z}_{12}\) may be
utilized for comparing the strength of phylogenetic signal among
datasets containing a different number of species, and even for
phenotypes obtained from species in different lineages, because their
phylogenetic non-independence and observed variation are taken into
account in the generation of the empirical sampling distribution via
permutation. \hfill\break

This study is not the first to compare \(\lambda\) and \(\kappa\) for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \(\kappa\) values are consistent with those found in
the simulations performed by Münkemüller et al.~(2012), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \(\kappa\) at
intermediate phylogenetic signal strengths was observed (see Fig. 2,
Münkemüller et al. 2012). Recent work by Molina-Venegas and Rodríguez
(2017) found that \(\kappa\) but not \(\lambda\) tended to inflate the
estimate of phylogenetic signal, leading to moderate type I and type II
biases, if polytomic chronograms were used. Their work more thoroughly
addressed previous observations of inflated \(\kappa\) for incompletely
resolved phylogenetic trees (Davies et al. 2012; Münkemüller et al.
2012). An interesting question is whether an inflated \(\kappa\) value
leads to an inflated \(Z_\kappa\) or does a tendency of a particular
tree to inflate estimates of \(\kappa\) also inflate the values in
random permutations of a test, in which case \(Z_\kappa\) is robust to
polytomies? We repeated the analyses in Figure 4, adjusting trees to
have 50\% collapsed nodes, per the technique of Molina-Venegas and
Rodríguez (2017), and found results were consistent (see Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \(\kappa\) as a descriptive statistic does not inflate
\(Z_\kappa\) as an effect size. Furthermore, because comparison of
effect sizes in a test is a comparison of locations of observed values
in their sampling distributions, which would shift concomitantly because
of this tendency, the \(Z_{12}\) test statistic in equation 4 appears to
be robust in spite of unresolved trees. \hfill\break
